Decoding Mortality: Trends and Causes in Canada between 2001-2022

### 2024-03-16 outputs/paper/paper.qmd
Update paper

library(tidyverse)
library(knitr)
library(rstanarm)
library(modelsummary)

@@

The remainder of this paper is structured as follows. In @sec-data, we showed the data and methodology utilized in this study. @sec-model discusses the model used to analyze the data

@@

R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data cleaning and visualization.

```{r}
#| label: Tbl-1
#| fig-cap: Top-ten causes of death in Canada in 2022
#| echo: false

@@

@@

```{r}
#| label: Fig-1
#| fig-cap: Annual number of deaths for the top-five causes in 2021, since 2001, for Canada
#| echo: false
#| warning: false
#| message: false

canada_top_five <-
  transformed_data |>
  filter(
    Year == 2022, 
    n == 22
  ) |>
  slice_max(order_by = desc(Ranking), n = 5) |>
  pull(Cause)

analysis_data <- cleaned_data |>
  filter(Cause %in% canada_top_five)

analysis_data |>
  ggplot(aes(x = Year, y = Death, color = Cause)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Year", y = "Annual number of deaths in Canada") +
  facet_wrap(vars(Cause), dir = "v", ncol = 1) +
  theme(legend.position = "none")

@@

```{r}
#| label: Tbl-2
#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
#| echo: false

datasummary(Death ~ Min + Mean + Max + SD + Var + N,
            fmt = 0,
            data = analysis_data)
```
@tbl-2 shows that the mean, 32,603, is different to the variance 673,243,267.

@@

```
# Model

The goal of our modelling strategy is twofold. 
Firstly,...
we would like to model the mortality in Canada.

Here we briefly describe the Bayesian analysis model used to investigate...

We model the incidences of death by one of the leading cause of deaths using a Poisson regression model and a negative Binomial regression model. Background details and diagnostics are included in [Appendix -@sec-model-details].

R packages rstanarm[@rstanarm] and modelsummary [@model] are used to build and analyze the models.

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

Since the residual of the number of deaths by a leading cause in a year does not follow a normal distribution, we consider using a generalized linear regression model. Since mortality occurrences are discrete events, we consider the Poisson regression model and its variant, negative binomial regression model, both of which relate explanatory variables to dependent variables representing counts of events (Reference: Poisson for aggregate data). Table 2 shows that the mean and variance are not the same. This indicates dispersion in the data, therefore negative binomial regression model may be more suitable.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

Table X presents the coefficients for various causes from both the Poisson and the Negative Binomial regression models.
The intercept represents the baseline of the model, in this case, Accidents (unintentional injuries). It shows the log count of deaths when all other causes of death are at this reference level, i.e. when no one other specific cause is considered.

The coefficients for specific causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of heart, and malignant neoplasms reflect the log-relative difference in the count of deaths relative to the reference level. Since we used a log link function, the coefficients represent effects on the log scale. Positive coefficients therefore indicate positive association between the cause and the number of deaths.

Table shows that among the top five leading causes of death in Canada, diseases of the heart with a coefficient of 1.454 and malignant neoplasms with a coefficient of approximately 1.805 significantly indicates a strong positive association in both models. The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model. Exponentiating the coefficient 1.454 to obtain `r round(exp(1.454),2)` suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries). Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is `r round(exp(1.805)-1, 2)` times higher Accidents (unintentional injuries). Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about `r round(1-exp(-0.070),2)` less. Both the Poisson model and the negative binomial model yielded similar coefficients.

This similarity suggests that both models are capturing similar relationship between the causes and the number of deaths. Comparing the Log Likelihood, The Negative Binomial model has a much higher (less negative) value -1050.577 than the Poisson model -16957.794. A higher log likelihood indicates a better model fit thus our analysis suggests the Negative Binomial model fits the data better. The other measures ELPD, LOOIC WAIC and RMSE all suggest that the negative binomial model fits the data better @sec-model-details

@@

```
# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

@@

Weaknesses and next steps should also be included.

>>>>

library(tidyverse)
library(knitr)
library(rstanarm)
library(modelsummary)
library(arrow)

@@

Death is a dreaded yet unavoidable event in life. As Benjamin Franklin famously stated, "In this world, nothing can be said to be certain, except death and taxes." Population mortality has been meticulously recorded over the past century, enabling the study of epidemiological trends, such as those observed in the recent pandemic. Crucially, analyzing region-specific mortality and identifying the leading causes of death inform public health decisions and guide the effective allocation of government resources to prevent and treat diseases. The availability of vast data sources unveils patterns and valuable information regarding causes of death.

This study investigates mortality in Canada from 2001 to 2022, utilizing the Canadian Vital Statistics Death Database by Statistics Canada. The focus is on identifying the risk factors associated with each cause of death. We employed a Bayesian approach, using both the Poisson regression model and the Negative Binomial regression model, to analyze the top five leading causes of mortality in Canada. The Bayesian model applies Bayesian inference to continually update the probability of a hypothesis as more data becomes available. The Poisson regression model is traditionally favored for count data but assumes equal mean and variance, an assumption that does not always hold in mortality data due to over-dispersion. The Negative Binomial regression model addresses this by accounting for over-dispersion, offering a more flexible approach for such data.

The top five leading causes of death identified are malignant neoplasm, diseases of the heart, chronic respiratory diseases, cerebrovascular disease, and unintentional accidents. Malignant neoplasm exhibits the strongest impact death counts, being 5 times more influential than accidents. Diseases of the heart account for 3 times more deaths than unintentional accidents. The findings advocate for prioritizing public health resources towards the screening, prevention, and treatment of these conditions.

The remainder of this paper is organized as follows. @sec-data discusses the data and measurement methods used in this study. @sec-model presents the model utilized in the analysis. @sec-result presents the results, and @sec-discussion discusses the implications, limitations, and proposes future directions for research.

@@

## Data Source

This study was inspired by the analysis of mortality in Alberta [@story], focusing on the leading causes of death, and extends the analysis to Canada, exploring the leading causes of death nationwide.  The dataset for this analysis is sourced from the open government portal of Statistics Canada, summarizing the causes of death and the number of individuals affected across all age groups from 2000 to 2022. This data originates from the Canadian Vital Statistics – Death Database (CVSD), an administrative survey that collects demographic and medical information, including the cause of death, from vital statistics registries in all provinces and territories, a practice ongoing since 1921.

## Data Measurement

The target population of the Canadian Vital Statistics – Death database (CVSD) includes deaths occurring in Canada among both Canadian residents and non-residents. As the survey operates on a census with a cross-sectional design, it does not employ sampling methods.

Data on registered deaths is submitted to Statistics Canada by each provincial and territorial Vital Statistics Registry. The death registration form comprises personal details provided to the funeral director by someone knowledgeable about the deceased and a medical certificate of cause of death, completed by the last attending medical professional or a coroner in cases necessitating an inquest or inquiry.

The data collected includes age, sex, marital status, residence and birthplace of the deceased, date of death, underlying cause of death according to the World Health Organization's International Statistical Classification of Diseases and Related Health Problems (ICD), province or territory of death occurrence, place of accident for non-transport accidental deaths, and autopsy details.

Statistics Canada conducts routine quality checks to ensure an error rate below 3%. However, potential biases may arise from the classification of death causes, adherence to the ICD, inclusion of resident data, and data transmission methods. Specifically, the shift in data collection for Canadian residents dying in the United States post-2009, the unavailability of Yukon data post-2017, and the transition to electronic data transmission via the National Routing System (NRS) could introduce biases. These factors may affect the survey's representativeness and accuracy, which are crucial for researchers to consider.

## Data Characteristics

The original dataset obtained from the Statistics Canada open government portal includes 2,268 observations across 18 variables, offering detailed demographic and geographical information on registered deaths. I aim to focus on five specific variables: 'Reference Period', 'Leading Causes of Death (ICD-10)', 'Characteristics', 'Ranking', and 'Number of Deaths'. The 'Reference Period' denotes the year of death registration (2000-2022), while 'Leading Causes of Death' categorizes the cause according to the International Statistical Classification of Diseases and Related Health Problems (ICD-10). 'Characteristics' encompasses both the 'Number of Deaths' and their 'Ranking', indicating either the total deaths or the disease's rank in a given year, respectively. A sample of the cleaned data can be found in @sec-data-details.

## Data Visualization

@tbl-1 highlights the top ten causes of death in 2022, with counts reflecting their frequency over 22 years. Except for COVID-19, all other causes have consistently been among the top ten annually. We selected the top five for detailed examination.

@fig-1 depicts the annual death trends for these top five causes. There has been a modest increase in deaths from Accidents (unintentional injuries) over two decades. Deaths due to Malignant Neoplasms have risen from 60,000 in the early 2000s to nearly 80,000 in 2023, while Cerebrovascular Diseases, Chronic Lower Respiratory Diseases, and Diseases of the Heart have maintained steady annual deaths at approximately 25,000, 20,000, and 60,000, respectively.

R [@citeR] was the language and environment used for the bulk of this analysis, alongside the tidyverse [@thereferencecanbewhatever], arrow[@arrow], knitr[@knitr] , ggplot2[@Wickham2016ggplot2] have been used in data downloading, cleaning and visualization.

@@

```
```{r}
#| label: fig-1
#| fig-cap: Annual number of deaths for the top-five causes in 2021, since 2001, for Canada
#| echo: false
#| warning: false
#| message: false

top_five <-
  transformed_data |>
  filter(
    Year == 2022, 
    n == 22
  ) |>
  slice_max(order_by = desc(Ranking), n = 5) |>
  pull(Cause)

# analysis_data <- cleaned_data |>
#   filter(Cause %in% canada_top_five)

cleaned_data |>
  filter(Cause %in% top_five)|>
  ggplot(aes(x = Year, y = Death, color = Cause)) +
  geom_line() +
  theme_minimal() +
  labs(x = "Year", y = "Annual number of deaths in Canada") +
  facet_wrap(vars(Cause), dir = "v", ncol = 1) +
  theme(legend.position = "none")

@@

```{r}
#| label: fig-2
#| fig-cap: Number of deaths from the top-five causes % of all deaths in 2021
#| warning: false
#| message: false

cleaned_data |>
  filter(Year == 2021)
  ggplot(top_five, aes(x="", y=total_deaths, fill=cause_of_death)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  theme_void() +
  labs(fill="Cause of Death")

@@

# Model {#sec-model}

The goal of our modeling strategy is to analyze mortality data for the leading causes of death in Canada over the past two decades, aiming to understand the relationship between disease and the number of deaths. We employ a Bayesian approach, assigning prior distributions to parameters based on prior knowledge and specifying a likelihood function based on the observed data. By applying Bayes' theorem, we update our prior beliefs to form posterior distributions of the parameters. Markov Chain Monte Carlo (MCMC) methods are utilized to approximate these posterior distributions, allowing us to integrate prior knowledge and interpret our findings within a probabilistic framework.

Since the residual of the number of deaths by a leading cause in a year does not follow a normal distribution, we consider using a generalized linear regression model. Since mortality occurrences are discrete events, we consider the Poisson regression model and its variant, negative binomial regression model, both of which relate explanatory variables to dependent variables representing counts of events. @tbl-2 shows that the mean, 32,603, is different from the variance, 673,243,267.  This indicates dispersion in the data, therefore negative binomial regression model may be more suitable.

```{r}
#| label: tbl-2
#| fig-cap: Summary Statistics of the number of yearly deaths, by cause, in Canada
#| echo: false

datasummary(Death ~ Min + Mean + Max + SD + Var + N,
            fmt = 0,
            data = analysis_data)
```

For model justification, we acknowledge that mortality occurrences are discrete events, making the Poisson and negative binomial regression models suitable for relating explanatory variables to event counts. The negative binomial model is particularly chosen for its capacity to handle overdispersion, a common feature in mortality data, under the assumption that the model's residuals are uncorrelated.

We model the incidences of death by one of the leading cause of deaths using a Poisson regression model and a negative Binomial regression model. Background details and diagnostics are included in [@sec-model-details].

R packages rstanarm[@rstanarm] and modelsummary [@model] are used to build and analyze the models. We use the default priors from `rstanarm`.

# Results {#sec-result}

@tbl-3 presents coefficients from both Poisson and Negative Binomial regression models for various causes of death. The intercept, representing the baseline of Accidents (unintentional injuries), indicates the log count of deaths when no other specific cause is considered. Coefficients for causes such as cerebrovascular diseases, chronic lower respiratory diseases, diseases of the heart, and malignant neoplasms reflect log-relative differences in death counts relative to Accidents. Positive coefficients indicate a positive association with the number of deaths.

@@

For instance, @tbl-3 shows that diseases of the heart have a coefficient of 1.454 in both models, suggesting the expected death count from this cause is approximately `r round(exp(1.454),2)`  times higher than from Accidents (unintentional injuries), after exponentiating the coefficient. Similarly, malignant neoplasms have a coefficient of approximately 1.805, indicating an expected count more than `r round(exp(1.805)-1, 2)` times higher than Accidents. Conversely, chronic lower respiratory diseases show a modest decrease in mortality compared to Accidents, with a coefficient of -0.070 in the Poisson model, indicating about `r round(1-exp(-0.070),2)` fewer deaths.

Table shows that among the top five leading causes of death in Canada, diseases of the heart with a coefficient of 1.454 and malignant neoplasms with a coefficient of approximately 1.805 significantly indicates a strong positive association in both models. The coefficient for Diseases of heart is 1.454 in both the Poisson model and the negative binomial model. Exponentiating the coefficient 1.454 to obtain suggests that the expected count of deaths due to Diseases of the heart is about `r round(exp(1.454) - 1,2)` times higher than the baseline category Accidents (unintentional injuries). Similarly, exponentiating the coefficient for malignant neoplasms to obtain `r round(exp(1.805),2)` shows the expected count of deaths is ` times higher Accidents (unintentional injuries). Conversely, chronic lower respiratory diseases are associated with a slight decrease in mortality compared to Accidents (unintentional injuries) with a coefficient of -0.070 in the Poisson model, although the impact is modest about  less.

Both models yield similar coefficients, suggesting they capture a similar relationship between causes and death counts. The Negative Binomial model, with a log likelihood of -1050.577, fits the data better than the Poisson model, which has a log likelihood of -16957.794. The negative binomial model's superior fit is further supported by a higher ELPD value compared to the Poisson model. Posterior predictive checks, trace plots, and Rhat plots assess MCMC convergence and the reliability of the Bayesian analysis, with details in @sec-model-details.

@@

# Discussion {#sec-discussion}

## Findings {#sec-first-point}

Our analysis over the past 20 years reveals that the expected count of deaths from malignant neoplasm, the leading cause of death, is approximately `r round(exp(1.805),2)` times higher than that from accidents (unintentional). Diseases of the heart, ranking second, account for `r round(exp(1.454) - 1,2)` times more deaths than accidents. This contrasts with global statistics from the WHO, highlighting the significance of allocating resources towards cancer prevention and treatment in Canada [@who]. The analysis of Alberta's data shows similarities with national trends, with the notable exception of organic dementia replacing accidents due to its correlation with aging, despite Alberta having the lowest percentage of senior population at 15.2% by 2023 [@aging]. This calls for further investigation into Alberta's health indicators and lifestyle factors.

## Ethical implications

The study ensures confidentiality and respectful data handling, particularly when dealing with rare causes of death. It raises concerns about potential disparities in resource access and safety, urging further research into occupational and transportation inequities. Sampling biases were minimized, but differential data access could lead to underrepresentation, notably for Yukon post-2017, potentially affecting policy benefits [@statcan]

@@

While the Poisson and Negative Binomial regression models demonstrate good fit, they assume independence of death counts, which may not reflect reality due to clustering by geographic and demographic factors. Additionally, we only considered one variable the leading cause of death, without considering rarer causes or potential confounders like socioeconomic factors. This constrains the ability of the model to capture the nuances of mortality and its drivers.

Future studies could aim to address these weaknesses by incorporating spatial and demographic variables to account for clustering effects. Future studies can broaden the scope to include more regions, socioeconomical variables, and their interactions. Longitudinal studies could offer insights into mortality trends and the long-term effects of environmental factors like climate change.
